# 3.3 Opening Data in jamovi {.unnumbered}

Before we can analyze data, we need to get it into jamovi and make sure it is set up correctly.

You can open existing datasets in jamovi, including files such as Excel spreadsheets and SPSS files.

When you open a file, take a moment to look through the data and ask:

-   Are the rows and columns organized the way I expect?
-   Are the variables labeled clearly?
-   Are the variable types set correctly?
-   Are there missing values or unusual entries?

That quick first look can save you a lot of problems later.

Alexander Swan has a video on [opening files in jamovi](https://youtu.be/nB85VyskK08).

```{r echo = FALSE, eval = knitr::is_html_output(excludes = "epub"), message = FALSE, warning = FALSE}
library(vembedr)
embed_url("https://youtu.be/nB85VyskK08")
```

### Opening SPSS (.sav) files in jamovi

If you are working with SPSS files (with a .sav file extension), jamovi can usually open them directly. This can be especially useful if you are working with older datasets or files shared by others.

Alexander Swan has an additional video on [opening .sav files in jamovi](https://www.youtube.com/watch?v=e-09OfLVs2U).

```{r echo = FALSE, eval = knitr::is_html_output(excludes = "epub"), message = FALSE, warning = FALSE}
library(vembedr)
embed_url("https://www.youtube.com/watch?v=e-09OfLVs2U")
```

### Entering data manually in jamovi

Sometimes you will need to enter data manually rather than opening an existing file.

If you are entering data manually:

-   make sure your variable names are clear
-   make sure each row is a case and each column is a variable
-   make sure variable types are set correctly
-   enter data consistently

This is also a good time to apply what we learned in Chapter 2 about good datasets and spreadsheets.

Alexander Swan has yet another video on [entering data manually in jamovi](https://youtu.be/g3a2hGwNZ7o).

```{r echo = FALSE, eval = knitr::is_html_output(excludes = "epub"), message = FALSE, warning = FALSE}
library(vembedr)
embed_url("https://youtu.be/g3a2hGwNZ7o")
```

## Annotating output in jamovi

Sometimes you may need to submit output files from jamovi with annotations. This can be helpful for showing your work, identifying key results, or explaining how you interpreted an analysis.

If you have not done this before, it is worth reviewing a short tutorial by Alexander Swan on [annotating output in jamovi](https://youtu.be/I6SZtH__ICA).

```{r echo = FALSE, eval = knitr::is_html_output(excludes = "epub"), message = FALSE, warning = FALSE}
library(vembedr)
embed_url("https://youtu.be/I6SZtH__ICA")
```
